### 2020-12-11 doc/manuscript/Dow_manuscript_Nature.Rmd
working on intro

As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[REFS] and annual CO~2~ sequestration[REFS].
However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually[REFS] but is the longest-lived carbon sink[REFS].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted *both leaf phenology* and woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
*The latter finding is confirmed on the centennial scale by tree-ring records across # forests.*
These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019,@zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests, implying that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.

@@

## (forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)
As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO2 from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon [@bonan_forests_2008]. However, their future under the current era of rapid change is uncertain [@friedlingstein_climatecarbon_2006].

Reducing this uncertainty requires better understanding of the drivers of tree growth [@bugmann_tree_2019]. Tree growth is sensitive to climatic factors like precipitation, temperature, and cloudiness in early growing season [@helcoski_growing_2019]; [@Ciais_europe-wide_2005]; [@Goulden_exchange_1996], as well as other physical factors including nitrogen content of soil, increasing atmospheric CO~2~, deposition of atmospheric pollutants in acid rain [@mathias_disentangling_2018], and invasion by pests or pathogens. Understanding how growth is influenced by these factors is essential for developing accurate models to inform forest managers and policy makers about the role of future forests in climate change mitigation efforts.

## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, an increase in average spring temperature has been impacting many natural systems including forests [@parmesan_globally_2003]; [@zhizong_fingerprinting_2006]. 
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*. For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010];@jeong_phenology_2011];@crabbe_extreme_2016]. The emergence of leaves signals the beginning of carbon sequestration or in other words, the growing season of a tree. It is believed that the longer the tree has leaves - or as the growing season is extended - the net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$; @churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016].)

However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019]. This implies that earlier leafout will lead to smaller increases in NEE than changes in other phenological traits like the maximum C uptake rate. This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE. A paper by @zani_increased_2020 found that earlier leaf out led to earlier leaf senescence within the deciduous trees in their study, reducing the ability of the trees to sequester significantly higher levels of C. Instead they posit that senescence is induced when seasonal carbon-sink capacity of the tree is reached, no matter how extended the potential growing season is. This result directly contradicts those previous predictions that warmer Spring temperatures and earlier leaf out will lead to drastically increased C sequestration.

## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
Earlier timing of spring leaf-up and initiation of carbon sequestration signals that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research on the characteristics of stem-growth during this shift has been done. In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns. *(but check into this.... I suspect there may be some that we're missing--perhaps auto dendro bands in Europe. Neil did say that there's little on early spring T)*.

Tree-ring records give some hints that warmer springs may increase annual growth ([@Helcoski_growing_2019]; probably some more), but tree-rings tell us nothing about the timing of that growth. On the other hand, earlier leaf development shows us how the growing season is advancing, but can't tell us about the total carbon sequestered, while neither can be used to map out intra-annual growth patterns.

Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests.A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals. A mixture of these three would produce the most complete picture of spring climate influence on forests.

## (We also know little about differences across species or functional groups, which matters because it will affect how different species fare under climate change )
*(Too much detail here? how about just keeping some variation of the first and last sentences)*
Leaf and wood phenology traits, along with their apparent relationships, are dissimilar among trees of different wood types (ring-porous and diffuse-porous; **D'Orangeville et al. **). Interspecific variation of this sort likely stems from evolutionary differences and adaptations to local climates where these trees can be found. Ring-porous trees have large xylem cells which are prone to embolism during the winter. These cells are repaired before leaf development in the Spring, corresponding with an earlier initiaiton of stem-growth. Diffuse-porous xylem cells are smaller, and less prone to embolism during the Winter, meaning they do not need to be repaired in order to restore hydraulic functionality before beginning the leaf development process. Results from the paper by D’Orangeville et al. (in review) have shown that ring-porous trees reach 25 and 50% of their total yearly growth during spring, while diffuse-porous trees reach these milestones in the summer, but both reached 75% total annual growth around the same time. If leaf development is sped up when spring temperatures are warmer, we should seek to understand how this shift influences stem-growth timing and other phenological variables.

## (Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings based on their differing growth phenology[*D'Orangeville*]. 
First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015]to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
*(We also look at leaf phenology.)*
Second, at these focal sites and *#* additional sites distributed *xxxx*, we analyze *#* tree-ring chronologies to understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale.

@@

**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: #-# days) and differed between ring- and diffuse-porous species (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`).**
Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 194 (July 14) at Harvard Forest.
Ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively). 
Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.

@@

>>>>

As the climate changes, warmer spring temperatures are causing earlier leaf-out[REFS] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[REFS] and annual CO~2~ sequestration[REFS].
However, little is known about how spring temperatures affect tree stem growth, which consumes a modest portion of the carbon fixed annually[REFS] but is the longest-lived carbon sink[REFS].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted *both leaf phenology* and woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
*The latter finding is confirmed on the centennial scale by tree-ring records across # forests.*
These findings are consistent with *~~most tree-ring studies in temperate deciduous forests showing that annual growth tends to be more sensitive to climatic conditions during the peak growing season than in the spring[REFS], with~~* eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests, implying that warming spring conditions are unlikely to impact temperate deciduous forest growth sufficiently to produce any substantial negative climate change feedback.

@@

## (ftemperate deciduous forests are critical for climate change regulation, so we need to understand better how they're responding to climate change)

(*this part needs to be re-written (very concisely, and focused on the temperate forest C sink and its uncertainty*)

As global atmospheric greenhouse gas levels are rising, and little meaningful policy is being implemented in response, our planet is expected to surpass the 2.5 degree C warming mark that the Paris accord pledged to avoid. The absence of human intervention increases our reliance on natural systems to soften the blow of climate change. Plant life, especially forests, make up a large portion of these natural systems, offering many valuable ecosystem services. For one, they house an astonishing 80% of terrestrial species and are capable of removing the greenhouse gas CO~2~ from the atmosphere and storing it in the trees or soil. Forests sequester about one-third of all anthropogenic CO~2~ and store ~45% of all terrestrial carbon[@bonan_forests_2008]. However, their future under the current era of rapid change is uncertain[@friedlingstein_climatecarbon_2006].

## (spring temperatures are rising, and this is affecting leaf phenology and carbon cycling)
In recent decades, warming temperatures have resulted in widespread phenological changes indicative of earlier arrival of spring[@parmesan_globally_2003] and (*often later autumn*[REFS].
In temperate deciduous forests, spring leaf-out has been occurring earlier[REFS], as has the beginning of the annual period of net CO~2~ sequestration[REFS].
The timing of autumn senescence is less straightforward (*need to do more lit review on this*).
In some regions earlier leaf out has been matched or exceeded by delays in leaf senescence *when springtime temperatures are above average(??)*[@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016]. (*is this just spring temperatures? Or autumn? how widespread is this?*)
Experimental evidence demonstrates that warmer spring temperatures lead to earlier leaf-out and earlier leaf senescence[@zohner_ongoing_2019], apparently because of sink-limitation to CO~2~ sequestration[@zani_increased_2020], whereas warmer summer and autumn temperatures delay leaf senescence[@zohner_ongoing_2019].
Whatever the mechanisms controlling autumn senescence, the observed pattern is that spring warming generally results in an increased growing season length, measured as the period over which trees have leaves[REFS] and the ecosystem is a net CO~2~ sink[REFS].

The expectation is that as growing season length increases, net annual carbon sequestered will increase as well (*i,e*, net ecosystem exchange, $NEE$)[@churkina_spatial_2005; @keenan_net_2014; @zhou_explaining_2016]. 
*This expectation is represented in models [@zani_increased_2020].*
However, recent evidence that leaf senescence is induced when the seasonal carbon-sink capacities of trees are reached suggests that warmer spring temperatures and earlier leaf-out will not lead to drastically increased C sequestration[@zani_increased_2020].

## (However, we don't know how warming springs affect stem-growth, which matters because the C allocated to stem growth will be retained in the ecosystem for decades to millenia )
Earlier timing of spring leaf-up and initiation of carbon sequestration signals that these trees are shifting their growth windows in response to increasing spring temperatures, but to date, very little research on the characteristics of stem-growth during this shift has been done. In fact, we know little about the stem-growth phenology of temperate deciduous species — let alone how these are affected by changes in spring temperature patterns (but see Ref. [D'Orangeville], *and we need to make sure we're not missing anything*).

Tree-ring records give some hints that warmer springs may increase annual growth (e.g., Refs.[@helcoski_growing_2019] [MORE]), but tree-rings tell us nothing about the timing of that growth. On the other hand, earlier leaf development shows us how the growing season is advancing, but can't tell us about the total carbon sequestered, while neither can be used to map out intra-annual growth patterns.

Thus, while it is increasingly clear that warming spring temperatures lengthen the growing season and modestly increase annual carbon sequestration, current methods do not allow firm conclusions on the future of forests. A combination of existing methods - tree cores and remote sensing of leaf-phenology traits - should be supplemented by dendrometer band data which allows for the modeling of intra-annual growth patterns in individuals. A mixture of these three would produce the most complete picture of spring climate influence on forests.

##(temporarily removed intro content)

(*Cameron, this is content --with associated refs--that I don't want to lose, but that needs to be worked into reworded text above*)
The effects of this spring warming are most apparent when examining leaf phenology shifts, *which have been extensively studied ([@way_tree_2011]; [@klosterman_later_2018]), leading to many critical observations*. 
For example, some regions have seen earlier leaf out in deciduous trees, which is often matched or exceeded by delays in leaf senescence timing, when springtime temperatures are above average [@ibanez_forecasting_2010;@jeong_phenology_2011;@crabbe_extreme_2016].

## (Here, we...)
Here, we characterize how early spring temperatures affect stem growth phenology, maximum daily growth rates, and annual growth of temperate deciduous trees, separated into ring- and diffuse-porous groupings, which differ in growth phenology[*D'Orangeville*]. 
First, using data from dendrometer bands measured throughout the growing season in two forests in the Eastern USA (n=`r n_tree_years` tree-years), we fit a growth model[@mcmahon_general_2015] to determine the days of year where 25, 50, and 75% annual growth was achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$), peak growing season length ($L_{pgs}= DOY_{75}$-$DOY_{25}$), maximum growth rates ($g_{max}$) and the DOY on which they occurred ($DOY_{g_{max}}$), and total annual growth ($\Delta DBH$; Fig. `r schematic_summary`). 
*(We also look at leaf phenology.)*
Second, at these focal sites and *#* additional sites distributed *xxxx*, we analyze *#* tree-ring chronologies to understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale.

@@

**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: #-# days) and differed between ring- and diffuse-porous species (Fig. `r growth_curves_all`, Table `r mean_growth_parameters_table`).**
Across all trees and years, $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 151 (June 1) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 194 (July 14) at Harvard Forest.
Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively). 
Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and <1 day longer for diffuse-porous species.

@@

However, more of the interannual variation in forest $NEE$ is explained by variation in maximum C uptake rates than by the length of the carbon uptake period [@zhou_explaining_2016; @fu_maximum_2019]. This idea agrees with patterns seen in tree cores, where peak growing season climate was more influential than spring climate [@helcoski_growing_2019] but many models still predict the extension of growing season will increase NEE.
